---
title: "PA2_NOAA_Storm.Rmd"
author: "Roger Fischer"
date: "April 22, 2015"
output: html_document
---

# TITLE
Title: Your document should have a title that briefly summarizes your data analysis

## SYNOPSIS
Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.


## ASSIGNEMENT
This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

Your data analysis must address the following questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.


## DATA PROCESSING
There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.


```{r}
## rm(list=ls())
## setwd("/Users/rogerfischer/datasciencecoursera/repdata/RepData_PeerAssessment2")

## sessionInfo()
## R version 3.1.2 (2014-10-31)
## Platform: x86_64-apple-darwin13.4.0 (64-bit)

## locale:
## [1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8

## attached base packages:
## [1] stats     graphics  grDevices utils     datasets  methods   base     

## other attached packages:
## [1] ggplot2_1.0.0 dplyr_0.4.1   knitr_1.9   

## Storm Data, 47Mb, zipped with bzip2 algorithm
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "StormData.csv.bz2", method="curl")
## read.csv can read zipped files directly
stormdata <- read.csv("StormData.csv.bz2")

## Early Exploration
dim(stormdata)
summary(stormdata$EVTYPE)
summary(stormdata$FATALITIES)
summary(stormdata$INJURIES)

```


## RESULTS
There should be a section titled Results in which your results are presented.


You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
